On Verification Patterns for Low-Level Systems via Modal Abstractions

### 2025-09-12 paper.tex
+

\subsection{Hybrid Logic, Dynamic Logic, and Nominals in Program Logics}
\label{sec:modalbackground}

\paragraph{Dynamic Logics as Program Logics}
We believe that for the kinds of informal reasoning and the sorts of data structures discussed in the previous section, ideas from modal logic are a promising approach to formalization and verification.
Broadly speaking modal logics incorporate \emph{modal operators}, which take as arguments a proposition expected to be true in another time~\cite{pnueli1977temporal}, place~\cite{gordon2019modal,goranko1996hierarchies,areces2001hybrid,gargov1993modal,murphy2004symmetric}, circumstance or point-of-view~\cite{hintikka1962knowledge,halpern1985guide}, and result in a proposition true in the \emph{current} time, place, or circumstance at which the truth of the use of the modal operator is being evaluated. Classic examples include modal necessity $\square P$ describing that $P$ is \emph{necessarily} true, $\mathsf{G}~P$ meaning $P$ is true \emph{globally} (i.e., forever from this time onwards in temporal logics), or $K_i(P)$ describing that a particular participant $i$ \emph{knows} that $P$ is true. The latter is an example of \emph{multimodal} logic, where there is an family of modalities (modal operators) parameterized by some dimension of interest (there, participants).

\emph{Dynamic logic}~\cite{pratt1976semantical}, a (multi)modal logic of weakest preconditions~\cite{dijkstra-75} which works with modalities of the form $[p](P)$, which states that \emph{in the current program state}, \emph{if} program $p$ is run then afterwards $P$ will hold (modulo non-termination).
This same idea is used to encode Hoare triples in the Iris program logic~\cite{jung2018iris}, using the same encoding as in Pratt's original presentation, where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$.
Iris is now one of the most successful foundational program logics, and its rich support for
extending the base logic with additional modalities~\cite{krebbers2017essence} means that
modal logics, and in particular dynamic logic, now play an outsized role in program verification ---
many challenging proofs of program correctness use a substructural dynamic logic containing
many additional modalities which interact with the weakest precondition modality.

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $\ell$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.

% \paragraph{Temporal Operators}
% The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
% due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
% distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.

\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
% This naming generally resembles
% the \emph{satisfaction} operator of 
% \emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. 
Because these named conditions strongly resemble the nominals of hybrid logic,
we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though not all of the examples we discuss are actually hybrid logics.

@@

% \paragraph{Virtual Memory Managers(VMM) with Multiple Address Spaces}
% \label{sec:backgroundonvmm}
% \todo{change the text}
% An intuition for the abstraction that \emph{contains} a page-table tree is called an \emph{address space} of a process whose virtual addresses to physical addresses. 
A typical OS manages multiple address spaces at once.
Each program is associated with a unique page table tree, which is stored in a specific hardware register (\texttt{cr3} in Figure \ref{fig:pagetables}). 
Using different mappings, which map only disjoint portions of physical memory in the map's codomain, 
is how the OS ensures memory isolation between processes.

\subsubsection{Virtualized Resources in Filesystems}
Another important computer resource computer resource that is \emph{virtualized} for access is disk blocks. 
Filesystems are software components that abstract disk access on behalf of normal programs
Similar to how VMMs using page table trees to fabricate more memory addresses than what actually exists, 
filesystems uses indexing maps for fabricating more \emph{virtual disk-page} addresses which are conventionally called \emph{disk-pages} 
to be distinguished from physical \emph{disk-block} addresses.
When a regular program accesses a file, it requests a specific logical offset from the start of that file's contents.
As with virtual memory, the filesystem may map adjacent logical blocks of a single file to physically disparate parts of the disk (or even
to different disks), though this translation mapping is implemented entirely in software (as opposed to the hardware cooperation that exists
for memory virtualization).

@@

% \begin{figure}[t]
%     \includegraphics[width=0.75\columnwidth]{cow_index.png}
%     \caption{A Filesystem called \textsf{Modfs} Indexing its Physical Pages for File Pages with CoW (copy-on-write) Tree}
%     \label{fig:cowindex}
% \end{figure}
% \paragraph{Examples to Different Policies on Navigating on Consistent Views of Disk}
% Like VMMs navigating between different address spaces, a \textsf{CoW} filesystem switches from either the old to the new root yields consistent views of the filesystem at different points in time, each commonly referred to as a \emph{snapshot}. A \textsf{CoW} filesystems treat the snapshot abstraction as VMMs treat the address space abstraction, container of the relevant resources. Like VMMs loading a process's memory mappings to be the current view of the memory by setting the control register to show the unique root address of the page-table tree of the process, the filesystem to atomically switch between consistent on-disk states, by updating the oldest so-called \emph{super} chunk at a fixed location on disk. Super chunks, as we see in Figure \ref{fig:cowindex} as ($\textsf{Super}_0$ and $\textsf{Super}_1$), are the first two blocks of the disk dedicated to keep the state of the file system which is composed of some meta-data on the filesystem, and most importantly to the root chunk of the filesystem index tree. 
% \begin{figure}
%     \includegraphics[width=\columnwidth]{journalling.png}
%     \caption{A Filesystem Utilizing \emph{Journalling} for Consistent Disk-Layout ~\cite{chajed2021gojournal}}
%     \label{fig:journal}
% \end{figure}
% \todo[inline,color=cyan]{This looks like it's verbatim copy-pasted from the ModFS draft; if this exact text
%   it used in this publication, it means we'd need to completely rewrite the relevant text in that draft.}
% Therefore, in case of a crash during the update to the \textsf{file "foo"}, we can reboot the file system from the previous persisted root, \textsf{root chunk}, via reading the $\textsf{Super}_0$. This means the whole disk consistency checks are unnecessary when recovering from a crash~\cite{bonwick2003zettabyte,hitz1994file}; recovery is immediate, as opposed to requiring hours or days~\cite{henson2006reducing,agrawal2009speedy} (or longer~\cite{henson2006chunkfs}) for recovering a large volume’s consistency from logs. 
% An important additional subtlety with this kind of filesystem is \emph{when} blocks in the reclaimed region can actually be reused: only \emph{after} the snapshot that made those blocks ``free'' has been completely written to disk. Allowing reuse before this could result in in-place updates: part of the new snapshot could attempt to store data in space used by the prior snapshot, allowing inconsistent on-disk structures.

While exceptions exist~\cite{bonwick2003zettabyte,rodeh2013btrfs,Rosenblum1992LFS,hitz1994file},
most filesystems will continue to store a given block of a given file in the same physical disk block for as long as the file exists.
These filesystems with in-place updates need to implement some form of \emph{journalling}~\cite{hagmann1987reimplementing,chutani1992episode} to ensure that a crash partway through an in-place update
can be rolled back using an undo journal that records the original contents before in-place modification (so original contents could be restored after a crash) 
or completed using a write-ahead journal that first writes new data in unused space before \emph{also} updating in place (so the journal entries could be re-applied after a crash). 
% \hrule
% In other words, unlike our ability to distinguish all resources (disk-pages and dis-kblocks) either being consistent under a snapshot or not in filesystems with \textsf{CoW}, with in-place updates, journalling mechanism does not allow such a claim. In other words, as we see in Figure \ref{fig:journal}, unlike \textsf{CoW}'s uniform persistency claim over disk resources under a snapshot as a resource container, elaborative bookkeeping mechanism of journalling leaves the filesystem treat disk resources in one of the resource groups in non-uniform fashion: disk-pages loaded into memory, updated disk-page not logged yet, updated and being logged or already logged disk-pages. Being able to treat the resources in a uniform fashion determines wwhether we need to consider the specification and proof of crash condition separately; instead a global never-violated invariant (not even temporarily broken) of the filesystem ensures the on-disk state is always consistent.
Because disk blocks associated with a file could now be in many states (on-disk, journaled but not updated in place, or both)
the specification of these systems is complex. And because some states are only possible after a system failure,
each line of filesystem code has two conceptual post-conditions: one referring to the successfully stored disk pages from a disk page container (e.g, a disk tree) , which can be uniquely identified by the last successful commit identifer (e.g., a transaction identifier or snapshot identifier), as the current view of the disk in the specification for continued execution, and one referring to the previous view of the disk in the specification for the stable state established in the event of a hardware failure immediately after that operation.

\subsubsection{Reference Counting Memory Reclamation}
In a typical configuration, allocating, deallocating and referencing memory locations for client code 
leaves the memory layout \emph{ideally} in a form of DAG (directed acyclic graph) with a unique root, which resembles the first stack pointer referencing the allocated memory. DAG abstraction instead of recursive decent tree allows non-duplication on the reference counts for the identical memory objects.
% To ease picturing the memory layout after a set of requests from the client code (for allocation, referencing etc.), we can think of the unique roots of DAGs as 
% \emph{super} block addresses -- this time multiple ones, not only two -- in Figure \ref{fig:cowindex}, 
% from which a path is created for a stack pointer reference to the allocated memory. 
%\todo[inline,color=cyan]{
%Why is this a DAG? Normally programs can easily create cycles, so the DAG bit requires a little more explanation.
%Do we mean that it is ``often'' the case that it's a DAG?}

@@

Reasoning about the correctness of such code requires reasoning about these reachability relations and reference counts, relative to specific
stack frames.

% \paragraph{Buffering} Buffering is a technique used in variety of systems (filesystem page caching, write buffer in multi-processor architectures, 
% translation-lookaside-buffers for translated addresses etc.) to delay the costly operation (committing file pages, flushing buffered writes from write-cache 
% to memory etc.) making updates consistent.
% \paragraph{Consistency of Reads of Buffered Writes} Consistency of reads of any value potentially buffered depends on the semantics of 
% the operation making them \emph{observable}. A typical architecture such as Intel-x86 considers the \emph{flush} of buffered writes 
% \emph{non-deterministically} --- i.e., at any time at any number of buffered writes. On the other hand, a typical filesystem can follow a certain 
% policy to commit the in-memory file pages to the disk.

\subsubsection{Weak Memory Models}
% I reworked this to be less general, and more about just weak memory, as thats all the survey section discusses.
Multi-core and multi-processor systems do not directly write every memory update back to memory (or even a coherent hardware cache)
individually, as this would be extremely slow. Instead, they effectively buffer reads and writes to and from memory
locally within each CPU core~\cite{owens2009better}\footnote{This is a layer below the better-known L1 cache, which is coherent across all CPUs in the system.}
so that operations can be partially reordered and batched to improve performance. For most code --- where the different CPUs are doing
independent work --- this is transparent to software. But fine-grained concurrent systems code which has multiple CPU cores
simultaneously coordinating access to shared memory must reason about these (effective) bufferings and reorderings, whose details vary
across CPU architectures.

Most of this coordination comes through reasoning about what \emph{will be true after local operations are published to memory},
or what \emph{will be true after remote operations become visible locally}~\cite{sieczkowski2015separation,dang2022compass,fsl}. Each of these flavors of contingent reasoning
depend on use of certain \emph{fence} or \emph{memory barrier} operations, which block execution of a CPU until either all incoming or outgoing
memory transactions are complete.

%\section{Resources in Systems Software}
%\label{sec:systemsoft}

%\subsection{Speculation in Consistency} \todo{speak about the weak-mem modalities in here.}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}

\section{Contingency Decomposition of a System}
In this section, we argue the position that \emph{modalities}  
should be a go-to approach when specifying and verifying low-level systems code.
We explain how the concept of a \emph{resource context} helps guide the design of new modalities
for verification of systems code, and
we justify our perspective by discussing prior systems that have used modalities for systems verification
successfully, arguing that they fit into the verification design pattern we articulate,
and explaining how this approach might apply to other systems verification challenges.

To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again, casting them in a certain way to bring out the relevance of our philosophy.

\paragraph{Shallow Modalities in the Iris Separation Logic}
\label{sec:modal_in_iris}
Classically modal operators are given direct semantics in terms of an underlying model, 
which for \textsf{Iris}~\cite{jung2018iris} would be a collection of \textsc{CMRA}s ---
an algebraic cousin of a BI-algebra~\cite{ohearn1999bunched}.
This is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{nakano-00}.
However in general, many modalities can be given proof-theoretic semantics directly within a logic. 
This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} 
logic itself), as well as for a wide range of other modalities in the 
literature~\cite{restall2002introduction}. These modalities are typically multi-modal, 
with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. 
In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ 
is typically replaced with substructural implication $\sepimp$.

@@

In addition, the modal definitions themselves may represent ownership of resources in a substructural setting~\cite{restall1993modalities,dovsen1992modal,kamide2002kripke}. 
This allows modalities to simultaneously represent ownership of certain resources used for interpreting the modality itself, as well as whatever resource ownership is implied by the modal argument.

\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\small
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

>>>>

\subsection{Hybrid Logic, Dynamic Logic, and Nominals in Program Logics}
\label{sec:modalbackground}
\paragraph{Dynamic Logics as Program Logics}
We believe that for the kinds of informal reasoning and the sorts of data structures discussed in the previous section, ideas from modal logic are a promising approach to formalization and verification.
Broadly speaking modal logics incorporate \emph{modal operators}, which take as arguments a proposition expected to be true in another time~\cite{pnueli1977temporal}, place~\cite{gordon2019modal,goranko1996hierarchies,areces2001hybrid,gargov1993modal,murphy2004symmetric}, circumstance or point-of-view~\cite{hintikka1962knowledge,halpern1985guide}, and result in a proposition true in the \emph{current} time, place, or circumstance at which the truth of the use of the modal operator is being evaluated. Classic examples include modal necessity $\square P$ describing that $P$ is \emph{necessarily} true, $\mathsf{G}~P$ meaning $P$ is true \emph{globally} (i.e., forever from this time onwards in temporal logics), or $K_i(P)$ describing that a particular participant $i$ \emph{knows} that $P$ is true. The latter is an example of \emph{multimodal} logic, where there is an family of modalities (modal operators) parameterized by some dimension of interest (there, participants).

\emph{Dynamic logic}~\cite{pratt1976semantical}, a (multi)modal logic of weakest preconditions~\cite{dijkstra-75} which works with modalities of the form $[p](P)$, which states that \emph{in the current program state}, \emph{if} program $p$ is run then afterwards $P$ will hold (modulo non-termination).
%This same idea is used to encode Hoare triples in the Iris program logic~\cite{jung2018iris}, using the same encoding as in Pratt's original presentation, where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$.
%Iris is now one of the most successful foundational program logics, and its rich support for
%extending the base logic with additional modalities~\cite{krebbers2017essence} means that
%modal logics, and in particular dynamic logic, now play an outsized role in program verification ---
%many challenging proofs of program correctness use a substructural dynamic logic containing
%many additional modalities which interact with the weakest precondition modality.

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $\ell$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
% \paragraph{Temporal Operators}
% The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
% due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
% distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
% This naming generally resembles
% the \emph{satisfaction} operator of 
% \emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. 
Because these named conditions strongly resemble the nominals of hybrid logic,
we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though not all of the examples we discuss are actually hybrid logics.

@@

% \paragraph{Virtual Memory Managers(VMM) with Multiple Address Spaces}
% \label{sec:backgroundonvmm}
% \todo{change the text}
% An intuition for the abstraction that \emph{contains} a page-table tree is called an \emph{address space} of a process whose virtual addresses to physical addresses. 
A typical OS manages multiple address spaces at once.
Each program is associated with a unique page table tree, which is stored in a specific hardware register (\texttt{cr3} in Figure \ref{fig:pagetables}). 
Using different mappings, which map only disjoint portions of physical memory in the map's codomain, 
is how the OS ensures memory isolation between processes. 
\subsubsection{Virtualized Resources in Filesystems}
Another important computer resource computer resource that is \emph{virtualized} for access is disk blocks. 
Filesystems are software components that abstract disk access on behalf of normal programs
Similar to how VMMs using page table trees to fabricate more memory addresses than what actually exists, 
filesystems uses indexing maps for fabricating more \emph{virtual disk-page} addresses which are conventionally called \emph{disk-pages} 
to be distinguished from physical \emph{disk-block} addresses.
When a regular program accesses a file, it requests a specific logical offset from the start of that file's contents.
As with virtual memory, the filesystem may map adjacent logical blocks of a single file to physically disparate parts of the disk (or even
to different disks), though this translation mapping is implemented entirely in software (as opposed to the hardware cooperation that exists
for memory virtualization).

@@

% \begin{figure}[t]
%     \includegraphics[width=0.75\columnwidth]{cow_index.png}
%     \caption{A Filesystem called \textsf{Modfs} Indexing its Physical Pages for File Pages with CoW (copy-on-write) Tree}
%     \label{fig:cowindex}
% \end{figure}
% \paragraph{Examples to Different Policies on Navigating on Consistent Views of Disk}
% Like VMMs navigating between different address spaces, a \textsf{CoW} filesystem switches from either the old to the new root yields consistent views of the filesystem at different points in time, each commonly referred to as a \emph{snapshot}. A \textsf{CoW} filesystems treat the snapshot abstraction as VMMs treat the address space abstraction, container of the relevant resources. Like VMMs loading a process's memory mappings to be the current view of the memory by setting the control register to show the unique root address of the page-table tree of the process, the filesystem to atomically switch between consistent on-disk states, by updating the oldest so-called \emph{super} chunk at a fixed location on disk. Super chunks, as we see in Figure \ref{fig:cowindex} as ($\textsf{Super}_0$ and $\textsf{Super}_1$), are the first two blocks of the disk dedicated to keep the state of the file system which is composed of some meta-data on the filesystem, and most importantly to the root chunk of the filesystem index tree. 
% \begin{figure}
%     \includegraphics[width=\columnwidth]{journalling.png}
%     \caption{A Filesystem Utilizing \emph{Journalling} for Consistent Disk-Layout ~\cite{chajed2021gojournal}}
%     \label{fig:journal}
% \end{figure}
% \todo[inline,color=cyan]{This looks like it's verbatim copy-pasted from the ModFS draft; if this exact text
%   it used in this publication, it means we'd need to completely rewrite the relevant text in that draft.}
% Therefore, in case of a crash during the update to the \textsf{file "foo"}, we can reboot the file system from the previous persisted root, \textsf{root chunk}, via reading the $\textsf{Super}_0$. This means the whole disk consistency checks are unnecessary when recovering from a crash~\cite{bonwick2003zettabyte,hitz1994file}; recovery is immediate, as opposed to requiring hours or days~\cite{henson2006reducing,agrawal2009speedy} (or longer~\cite{henson2006chunkfs}) for recovering a large volume’s consistency from logs. 
% An important additional subtlety with this kind of filesystem is \emph{when} blocks in the reclaimed region can actually be reused: only \emph{after} the snapshot that made those blocks ``free'' has been completely written to disk. Allowing reuse before this could result in in-place updates: part of the new snapshot could attempt to store data in space used by the prior snapshot, allowing inconsistent on-disk structures.
While exceptions exist~\cite{bonwick2003zettabyte,rodeh2013btrfs,Rosenblum1992LFS,hitz1994file},
most filesystems will continue to store a given block of a given file in the same physical disk block for as long as the file exists.
These filesystems with in-place updates need to implement some form of \emph{journalling}~\cite{hagmann1987reimplementing,chutani1992episode} to ensure that a crash partway through an in-place update
can be rolled back using an undo journal that records the original contents before in-place modification (so original contents could be restored after a crash) 
or completed using a write-ahead journal that first writes new data in unused space before \emph{also} updating in place (so the journal entries could be re-applied after a crash). 
% \hrule
% In other words, unlike our ability to distinguish all resources (disk-pages and dis-kblocks) either being consistent under a snapshot or not in filesystems with \textsf{CoW}, with in-place updates, journalling mechanism does not allow such a claim. In other words, as we see in Figure \ref{fig:journal}, unlike \textsf{CoW}'s uniform persistency claim over disk resources under a snapshot as a resource container, elaborative bookkeeping mechanism of journalling leaves the filesystem treat disk resources in one of the resource groups in non-uniform fashion: disk-pages loaded into memory, updated disk-page not logged yet, updated and being logged or already logged disk-pages. Being able to treat the resources in a uniform fashion determines wwhether we need to consider the specification and proof of crash condition separately; instead a global never-violated invariant (not even temporarily broken) of the filesystem ensures the on-disk state is always consistent.
Because disk blocks associated with a file could now be in many states (on-disk, journaled but not updated in place, or both)
the specification of these systems is complex. And because some states are only possible after a system failure,
each line of filesystem code has two conceptual post-conditions: one referring to the successfully stored disk pages from a disk page container (e.g, a disk tree) , which can be uniquely identified by the last successful commit identifer (e.g., a transaction identifier or snapshot identifier), as the current view of the disk in the specification for continued execution, and one referring to the previous view of the disk in the specification for the stable state established in the event of a hardware failure immediately after that operation.
\subsubsection{Reference Counting Memory Reclamation}
In a typical configuration, allocating, deallocating and referencing memory locations for client code 
leaves the memory layout \emph{ideally} in a form of DAG (directed acyclic graph) with a unique root, which resembles the first stack pointer referencing the allocated memory. DAG abstraction instead of recursive decent tree allows non-duplication on the reference counts for the identical memory objects.
% To ease picturing the memory layout after a set of requests from the client code (for allocation, referencing etc.), we can think of the unique roots of DAGs as 
% \emph{super} block addresses -- this time multiple ones, not only two -- in Figure \ref{fig:cowindex}, 
% from which a path is created for a stack pointer reference to the allocated memory. 
%\todo[inline,color=cyan]{
%Why is this a DAG? Normally programs can easily create cycles, so the DAG bit requires a little more explanation.
%Do we mean that it is ``often'' the case that it's a DAG?}

@@

Reasoning about the correctness of such code requires reasoning about these reachability relations and reference counts, relative to specific
stack frames.
% \paragraph{Buffering} Buffering is a technique used in variety of systems (filesystem page caching, write buffer in multi-processor architectures, 
% translation-lookaside-buffers for translated addresses etc.) to delay the costly operation (committing file pages, flushing buffered writes from write-cache 
% to memory etc.) making updates consistent.
% \paragraph{Consistency of Reads of Buffered Writes} Consistency of reads of any value potentially buffered depends on the semantics of 
% the operation making them \emph{observable}. A typical architecture such as Intel-x86 considers the \emph{flush} of buffered writes 
% \emph{non-deterministically} --- i.e., at any time at any number of buffered writes. On the other hand, a typical filesystem can follow a certain 
% policy to commit the in-memory file pages to the disk.   
\subsubsection{Weak Memory Models}
% I reworked this to be less general, and more about just weak memory, as thats all the survey section discusses.
Multi-core and multi-processor systems do not directly write every memory update back to memory (or even a coherent hardware cache)
individually, as this would be extremely slow. Instead, they effectively buffer reads and writes to and from memory
locally within each CPU core~\cite{owens2009better}\footnote{This is a layer below the better-known L1 cache, which is coherent across all CPUs in the system.}
so that operations can be partially reordered and batched to improve performance. For most code --- where the different CPUs are doing
independent work --- this is transparent to software. But fine-grained concurrent systems code which has multiple CPU cores
simultaneously coordinating access to shared memory must reason about these (effective) bufferings and reorderings, whose details vary
across CPU architectures.

Most of this coordination comes through reasoning about what \emph{will be true after local operations are published to memory},
or what \emph{will be true after remote operations become visible locally}~\cite{sieczkowski2015separation,dang2022compass,fsl}. Each of these flavors of contingent reasoning
depend on use of certain \emph{fence} or \emph{memory barrier} operations, which block execution of a CPU until either all incoming or outgoing
memory transactions are complete.
%\section{Resources in Systems Software}
%\label{sec:systemsoft}
%\subsection{Speculation in Consistency} \todo{speak about the weak-mem modalities in here.}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}
\section{Contingency Decomposition of a System}
In this section, we argue the position that \emph{modalities}  
should be a go-to approach when specifying and verifying low-level systems code.
We explain how the concept of a \emph{resource context} helps guide the design of new modalities
for verification of systems code, and
we justify our perspective by discussing prior systems that have used modalities for systems verification
successfully, arguing that they fit into the verification design pattern we articulate,
and explaining how this approach might apply to other systems verification challenges.

To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again, casting them in a certain way to bring out the relevance of our philosophy.
\subsection{Encoding Modalities in Program Logics}

\paragraph{Encoding Hoare Triples} With Pratt's original representation, we can encode where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$. However, this modal representation requires to be given a semantics which is in terms of an underlying model.

On the other hand the underlying model for \textsf{Iris}~\cite{jung2018iris} would be a collection of \textsc{CMRA}s ---
an algebraic cousin of a BI-algebra~\cite{ohearn1999bunched}.
This is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{nakano-00}.
However in general, many modalities can be given proof-theoretic semantics directly within a logic. 
This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} 
logic itself) , as well as for a wide range of other modalities in the 
literature~\cite{restall2002introduction}. These modalities are typically multi-modal, 
with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. 
\paragraph{Shallow Modalities in Iris Separation Logic}
\label{sec:modal_in_iris}
In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ 
is typically replaced with substructural implication $\sepimp$.

@@

In addition, the modal definitions themselves may represent ownership of resources in a substructural setting~\cite{restall1993modalities,dovsen1992modal,kamide2002kripke}. 
This allows modalities to simultaneously represent ownership of certain resources used for interpreting the modality itself, as well as whatever resource ownership is implied by the modal argument.
\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\small
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}
